---
title: "Activity prediction using sensor data"
author: "Dominic Tabeta"
date: "May 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

This project uses sensor data taken from activity sensors labelled with activity types and trains a machine learning model that predicts activity types using this sensor data as predictors. We also carry out cross validation within the training set.

## Data Source
The source for the training data set: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The source for the test data set: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project originated from here:
http://groupware.les.inf.puc-rio.br/har

## Initial Library setup

```{r initialsetup,echo=TRUE}
library(caret)
library(rpart)
library(randomForest)
library(rpart.plot)
library(rattle)
```

## Data loading

```{r loadingandpreprocessing}
setwd("~/datascience/pmlproject")
trainurl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testurl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainfile <- "../data/pml-training.csv"

if (!file.exists(trainfile)) {
        download.file(trainurl,destfile=trainfile,quiet=TRUE,mode="wb")
}
testfile <- "../data/pml-testing.csv"
if (!file.exists(testfile)) {
        download.file(testurl,destfile=testfile,quiet=TRUE,mode="wb")
}

traindata <- read.csv((trainfile), na.strings=c("NA","#DIV/0!",""))
testdata <- read.csv((testfile), na.strings=c("NA","#DIV/0!",""))
```

## Reproducibility

FOr reproducibility we set the random number seed here
``` {r setseed}
set.seed(123)
```

## Partitioning the training data set

For assessment of the machine learning models prior to application on the final test set I partition the training set here. The training-test split selected is 80-20.
``` {r partition}
selecttrain<-createDataPartition(y=traindata$classe,p=0.8,list=FALSE)
mytrain<-traindata[selecttrain,]
mytest<-traindata[-selecttrain,]
```

## Data Processing/Cleaning
We remove the index column and any variables that have zero variance from the mytrain and mytest datasets. Also we remove the columns that have over 97% missing values.
``` {r cleaning}
nzv <- nearZeroVar(mytrain, saveMetrics=TRUE)
mytrainclean<-mytrain[,row.names(nzv[nzv$percentUnique>0&nzv$percentUnique<100,])]


nacheck<-apply(mytrainclean,2,is.na)

nasum<-apply(nacheck,2,sum)
naprop<-nasum/nrow(mytrainclean)

mytrainclean<-mytrainclean[,names(naprop[naprop<.95])]

mytestclean<-mytest[,colnames(mytrainclean)]
testfinal<-testdata[,colnames(mytrainclean[-59])]

dim(mytrainclean)
dim(mytestclean)
dim(testfinal)
```

## Prediction model using Decision Trees (rpart)
``` {r rpartmodel}
model1 <- rpart(mytrain$classe ~ ., data=mytrainclean[-59], method="class")
fancyRpartPlot(model1)
prediction1 <- predict(model1, mytestclean, type = "class")
confusionMatrix(prediction1, mytest$classe)
```

## Prediction model using Random Forest
``` {r randomForest}
model2 <- randomForest(mytrain$classe ~. , data=mytrainclean[-59])
prediction2 <- predict(model2, mytestclean, type = "class")
confusionMatrix(prediction2, mytest$classe)
```

## Applying predictor models to the given test data

```{r finaltest}
prediction3<-predict(model2,cbind(testfinal,classe=rep("A",20)),type="class")
confusionMatrix(prediction3,testfinal$classe)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
